---
title: "PA1_templateRev"
author: "pmp"
date: "Sunday, December 07, 2014"
output: html_document
---

## REPRODUCIBLE RESEARCH - Peer Assessment Project 1
  

## Introduction

As stated in the introduction to this assignment, *[the] assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day*.

## Loading & Processing Data  

The data-set for the assignment was supplied as a zipped file called *Activtiy monitoring data.zip* and the data stored in a variable named "activity".    

```{r data loading}
filename <- "./activity.csv"
activity <- read.csv(filename)
```

This project rises  4 questions regarding the activity data-set whose answers form the core of this report:  

* What is mean total number of steps taken per day?
* What is the average daily activity pattern?
* Missing values and the effect of imputing for missing values
* Are there differences in activity patterns between weekdays and weekends?


## Data Explaration  

```{r summary data} 
summary(activity)
```

The data-set contains three variables representing:

.steps: Number of steps recorded in a 5-minute interval (missing values are coded as NA)  


.date: The date on which the measurement was taken in YYYY-MM-DD format  


.interval: Identifier for the 5-minute interval in which measurement was taken    


# Data Processing  

the following R libraries were necessary for this report  

```{r libraries}
library (data.table)
library( plyr)
library(lattice)
library(ggplot2)
library(grid)
library(gridExtra)
```


## Removing NA values  
Histograms representing the distribution of number of steps per interval for each of the 61 days, show a similar range in the frequency distribution values, as well as, in the number of steps per interval.   

```{r histo steps per interval per day}
histogram(~steps|date, data=activity,layout=c(7,9), par.strip.text=list(cex=.75))
```

The differences among steps per day, however, are more apparent if the log base 10 of the number of steps per interval are used:  

```{r histo log steps per interaval per day}
histogram(~log10(steps)|date, data=activity,layout=c(7,9), par.strip.text=list(cex=.75))
```

One can notice very clearly that most of the missing data (NA values) corresponds not so much with isolated measurements but with the absence of recorded measurements in nine of the 61 days monitored.  

## Data set without missing values  
```{r remove missing values}
activity1 <- activity[complete.cases(activity),]
summary (activity1)
```
---
---

# Analysis and Data Query:  
---

## What is the mean total number of steps taken per day?  

The mean and median values for the number of steps per day were calculated from the sum of steps for all intervals in a given day, its distribution of values shown below:  
```{r mean and median and histo per day}
steps.day <- aggregate(activity1$steps,by=list(activity1$date),sum)
mean.activity <- mean(steps.day$x)
median.activity <- median(steps.day$x)
hist(steps.day$x, breaks=30, col="orange", main = "Number of steps per day", xlab="number of steps", ylab="frequencey")
```

For comparison, the mean and median steps per interval, were also calculated  
```{r mean and median per interval}
mean.steps <- mean (activity1$steps,na.rm=TRUE)
sd.activity <- sd(steps.day$x)
median.steps <- median(activity1$steps,na.rm=TRUE)

df<- data.frame(mean.activity,sd.activity,median.activity,mean.steps,median.steps)
colnames(df) <-c("- mean per day","+/- sdev per day","- median per day", "- mean per interval", "-median per interval-")
```

as shown in the table below  
```{r meanmedian table}
df
```
 


---

## What is the average daily activity pattern?  

A time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis) shows the following pattern:  

```{r time series plot}
daily.activity <- aggregate(steps~interval, data=activity, mean)
daily.activity <- mutate(daily.activity, 
                interval=(interval/100))
                # interval=((interval/100)+100*(((interval/100)%%1)/60)))

x<- which.max(daily.activity$steps)
plot (daily.activity$interval, daily.activity$steps, type="l", 
      main =" daily activity",xlab="hours", ylab="average number of steps",
      xaxt='n')
axis(1,at = c(0,6,12,18,24))
```

With the 104th interval, corresponding to the 8:35am interval, showing the highest mean step activity from October to November 2012 for the given monitored individual  

```{r hieghest activity}
daily.activity[x,]
```

---  
---
#  Imputing missing values  

Imputation is a process used to generate values to replace NA missing values in a data-set. The use of imputation depends on the nature of the data [Missing data imputation](http://www.stat.columbia.edu/~gelman/arm/missing.pdf).  

```{r number of NAs}
number.na <- sum(is.na(activity$steps))
number.na
```
In this particular data set the total number of NA is relatively high at `r number.na` NAs. As an example of the imputation process, and how the methods used might or might not introduce changes in the general statistics of the data, I will use two imputation methods, one based on the mean and a second one based on a random selection of values to replace missing values.The techniques are borrowed from [Thomas Leeper](http://thomasleeper.com/Rcourse/Tutorials/NAhandling.html).   

A new imputed data-set named "activity2" is based on mean value substitutions: 

```{r mean value imputation}
activity2 <- activity
activity2$steps[is.na(activity2$steps)] <- mean (activity2$steps, na.rm=TRUE)
```

and a new data-set named "activity3" is based on random replacement imputation  

```{r Random value replacement imputation}
activity3 <- activity
samples <- activity3$steps[!is.na(activity3$steps)]
activity3$steps[is.na(activity3$steps)] <- sample(activity3$steps[!is.na(activity3$steps)],
                                            sum(is.na(activity3$steps)), TRUE)
```

### Comparisons of the two imputatation methods vs the original data  

First in terms of the distributions of the step numbers per interval per day:  
```{r comparison histogram 1}
histogram(~log10(steps)|date, data=activity,layout=c(7,9), main="Original Dataset", par.strip.text=list(cex=.75))
histogram(~log10(steps)|date, data=activity3,layout=c(7,9), main="Random Value Imputed Dataset", par.strip.text=list(cex=.75))
```

We can noticed that the random selection values method (RSVM) fills in those days with missing data with similarly distributed frequency distributions as those observed for other days, suggesting that the RSV imputation method recreates the frequency distribution observed in the original data fairly well.    
The use of the mean value imputation, on the other hand, does alter the distribution patterns for those days with missing values but not the overall means:  
```{r comparison histogram 2}
histogram(~log10(steps)|date, data=activity2,layout=c(7,9), main="Mean Imputed Dataset", par.strip.text=list(cex=.75))
```
---
Second, comparing the distributions of steps per day between the original data set and the RSV imputation is shown in the plot below:  

```{r comparison histogam per day}
steps.day3 <- aggregate(activity3$steps,by=list(activity3$date),sum)

steps.day$method <- "none"
steps.day3$method <- "RVS imputed"

comparison <- rbind(steps.day,steps.day3)
ggplot(comparison, aes((x), fill=method)) + geom_density(alpha=.2)
```

The plot shows that, as would be expected, the total number of steps is higher in the RSV imputed data set compared to the original data set. As far as the RSV imputed data set being a valid correction of missing values is in part supported by the distributions being very similar between the two data sets, although the RSV imputation has smoothed the original distribution eliminating some of the data-set's non-normalrandom characteristics.  

The R summary of the RSV imputed data set (activity3) is compared to the summary of the NA removed original data set (activity1)  
```{r comparison imputet data set}
summary (activity1)
summary (activity3)
```
and the table below compares the mean +/- standard deviation of daily average steps between the imputed and original data set:  
```{r mean and std table}
mean.activity3 <- mean(steps.day3$x)
sd.activity3<- sd(steps.day3$x)
median.activity3 <- median(steps.day3$x)
df2<- data.frame(mean.activity,sd.activity,mean.activity3,sd.activity3)
colnames(df2) <-c("- mean per day","+/- sdev per day","- imputetd mean per day","+/- sdev per day -")
df2
```


## Are there differences in activity patterns between weekdays and weekends?  

For the next query we are using the RVSM imputed data set for the analysis. 

Before embarking in the presumption that there is a difference between weekday and weekend activities one needs to compare not only weekdays  versus weekends as a whole but each week day versus each other:  
```{r processing time data}
activity3$date <- as.POSIXlt(activity3$date)
activity3$day <-weekdays(as.Date(activity3$date))
activity3$day <- factor(activity3$day, levels= c("Monday", 
                "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

weekly.steps <- by(activity3$steps, as.factor(activity3$day), mean)
```

The resulting plot  bellow indicates that steps per five minute interval increases in the weekends and in the middle of the week. Without knowing the particular activities of the individual monitored, the elevated average number of steps per interval on a Wednesday could also be due to an increased stepping tempo. It is also evident that the "weekend" trend starts on Friday, a day that needs to be included as part of the weekend:  

```{r mean steps per interval par week day}
plot(weekly.steps, main = "average number of steps per interval by day of the week", 
     ylab="mean number of steps", xlab="day of the week", type='l', xaxt='n', cex=0.85)
axis(1, at = 1:7, labels=c("Monday","Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"), cex.axis=0.75)
abline(h=mean(weekly.steps), col=3, lty=2)
```

A better insight of the average weekly 24 hour activity pattern is shown in the next plot generated comparing the average day of the week 24 hour activity.  

### Day of the week datasets:  
```{r day of the week data sets}
Monday.steps <- subset( activity, activity3$day =="Monday")
Monday.activity <- aggregate(steps~interval, data=Monday.steps, mean)
Monday.activity <- mutate(Monday.activity,interval=(interval/100))

Tuesday.steps <- subset( activity, activity3$day =="Tuesday")
Tuesday.activity <- aggregate(steps~interval, data=Tuesday.steps, mean)
Tuesday.activity <- mutate(Tuesday.activity,interval=(interval/100))

Wednesday.steps <- subset( activity, activity3$day =="Wednesday")
Wednesday.activity <- aggregate(steps~interval, data=Wednesday.steps, mean)
Wednesday.activity <- mutate(Wednesday.activity,interval=(interval/100))

Thursday.steps <- subset( activity, activity3$day =="Thursday")
Thursday.activity <- aggregate(steps~interval, data=Thursday.steps, mean)
Thursday.activity <- mutate(Thursday.activity,interval=(interval/100))

Friday.steps <- subset( activity, activity3$day =="Friday")
Friday.activity <- aggregate(steps~interval, data=Friday.steps, mean)
Friday.activity <- mutate(Friday.activity,interval=(interval/100))

Saturday.steps <- subset( activity, activity3$day =="Saturday")
Saturday.activity <- aggregate(steps~interval, data=Saturday.steps, mean)
Saturday.activity <- mutate(Saturday.activity,interval=(interval/100))

Sunday.steps <- subset( activity, activity3$day =="Sunday")
Sunday.activity <- aggregate(steps~interval, data=Sunday.steps, mean)
Sunday.activity <- mutate(Sunday.activity,interval=(interval/100))
```


###  Plot of Day of the Week activities  
Orange corresponds to the standard weekdays (Monday through Thursday) and blue corresponds to the weekend days (Friday through Sunday). The y scale is fixed between 0 and 350 steps.  

```{r compariative time series per week day}
plot.new()

par (mar=c(1,1,1,1))
par(mfrow = c(5, 2), cex=.5)

plot (Monday.activity$interval, Monday.activity$steps, type="l", ylim=c(0,350), col="orange",
      main =" Monday activity",xlab="hours", ylab="average number of steps",
      xaxt='n')
axis(1,at = c(0,6,12,18,24))
plot (Friday.activity$interval, Friday.activity$steps, type="l", ylim=c(0,350), col="blue",
      main =" Friday activity",xlab="hours", ylab="average number of steps",
      xaxt='n') 
axis(1,at = c(0,6,12,18,24))
plot (Tuesday.activity$interval, Tuesday.activity$steps, type="l", ylim=c(0,350),col="orange",
      main ="Tuesday activity",xlab="hours", ylab="average number of steps",
      xaxt='n')
axis(1,at = c(0,6,12,18,24))
plot (Saturday.activity$interval, Saturday.activity$steps, type="l", ylim=c(0,350), col="blue",
      main =" Saturday activity",xlab="hours", ylab="average number of steps",
      xaxt='n')
axis(1,at = c(0,6,12,18,24))
plot (Wednesday.activity$interval, Wednesday.activity$steps, type="l", ylim=c(0,350),col="orange",
      main ="Wednesday activity",xlab="hours", ylab="average number of steps",
      xaxt='n')
axis(1,at = c(0,6,12,18,24))
plot (Sunday.activity$interval, Sunday.activity$steps, type="l", ylim=c(0,350), col="blue",
      main =" Sunday activity",xlab="hours", ylab="average number of steps",
      xaxt='n')
axis(1,at = c(0,6,12,18,24))
plot (Thursday.activity$interval, Thursday.activity$steps, type="l", ylim=c(0,350),col="orange",
      main ="Thursday activity",xlab="hours", ylab="average number of steps",
      xaxt='n')
axis(1,at = c(0,6,12,18,24))
```

*Three main patterns are evident from the data:  
-Somewhere between 8am and 9am this individual performed the longest consecutive walk/run except for Sunday
-Thursday is the day with the least number of steps but with two consecutive long walks/runs.
-The weekend days have the highest overall daily activity.

### Are the differences between Weekdays and Weekend activity significant?  


```{r comparison means per day data sets}
selection <- c("Monday","Tuesday","Thursday")
Weekday.steps <- subset( activity, activity3$day == selection)
Weekday.mean.steps <- aggregate(Weekday.steps$steps,by=list(Weekday.steps$date),sum)
mean.weekday <- mean(Weekday.steps$steps, na.rm=TRUE)

mean.Weekday.day <- mean(aggregate(Weekday.steps$steps,by=list(Weekday.steps$date),sum)$x, na.rm=TRUE)
sd.Weekday.day <- sd(aggregate(Weekday.steps$steps,by=list(Weekday.steps$date),sum)$x, na.rm=TRUE)

selection2 <- c("Friday","Saturday","Sunday")
Weekend.steps <- subset( activity, activity3$day == selection2)
Weekend.mean.steps <- aggregate(Weekend.steps$steps,by=list(Weekend.steps$date),sum)
mean.weekend <- mean(Weekend.steps$steps, na.rm=TRUE)

mean.Weekend.day <- mean(aggregate(Weekend.steps$steps,by=list(Weekend.steps$date),sum)$x, na.rm=TRUE)
sd.Weekend.day <- sd(aggregate(Weekend.steps$steps,by=list(Weekend.steps$date),sum)$x, na.rm=TRUE)
```

The next plot compares the average daily values for the weekdays vs the weekends.    

```{r comparison weekday vs weekend}
p1<- qplot(Group.1, x, data=Weekday.mean.steps, geom="boxplot", ylim=c(0,7000), alpha=.5, main="weekdays" )
p1 <- p1 + theme(axis.title.x=element_blank()) + geom_hline(aes(yintercept=mean.Weekday.day))

p2 <-qplot(Group.1, x, data=Weekend.mean.steps, geom="boxplot", ylim=c(0,7000), alpha=.5, main="weekend" )
p2 <- p2 + theme(axis.title.x=element_blank()) + geom_hline(aes(yintercept=mean.Weekend.day))
grid.arrange(p1,p2,ncol=2,main="comparison between weekdays and weekend log10 of steps")
```

One can see a definite increase in the mean of the values, as well as, a bias towards larger numbers of steps during the weekend.  

Furthermore, the mean per interval number of steps like the mean per daily steps is increased during the weekend, as shown in the table below: "Speedy Gonzales on the weekends!"  

```{r weekday weekend mean and sd table}
df3<- data.frame(mean.weekday,mean.weekend)
colnames(df3) <-c("- mean per interval per weekday","- mean per interval per weekend -")
df3

df4<- data.frame(mean.Weekday.day,sd.Weekday.day,mean.Weekend.day,sd.Weekend.day)
colnames(df4) <-c("- mean per day per weekday"," +/- SD","- mean per day per weekend"," +/- SD -")
df4
```

---
---

### END
 